---
title: "Completely Randomized Design (CRD)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Completely Randomized Design (CRD)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## When to Use

The **Completely Randomized Design (CRD)** is the simplest experimental design.
Use it when:

- There is **one treatment factor** with two or more levels.
- Experimental units are **homogeneous** (no known sources of variability to block).
- Treatments are **randomly assigned** to units without restriction.

CRD is common in laboratory experiments and greenhouse trials where environmental
conditions are uniform.

## The Design

In a CRD, each treatment is assigned to experimental units purely at random.
The statistical model is:

$$Y_{ij} = \mu + \tau_i + \varepsilon_{ij}$$

where $\mu$ is the overall mean, $\tau_i$ is the effect of treatment $i$,
and $\varepsilon_{ij} \sim N(0, \sigma^2)$.

## Data

We use a dataset comparing the yield of three squash varieties (Acorn, Butternut,
Spaghetti) with three replicates each.

```{r data}
library(agrideshr)
data(crd_data)
str(crd_data)
crd_data
```

## Exploratory Visualization

```{r eda}
library(ggplot2)

ggplot(crd_data, aes(x = Variety, y = Yield, fill = Variety)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16) +
  geom_jitter(width = 0.1, size = 2) +
  theme_bw() +
  labs(title = "Yield by Squash Variety", y = "Yield", x = "Variety") +
  theme(legend.position = "none")
```

Descriptive statistics by variety:

```{r desc}
library(dplyr)

crd_data %>%
  group_by(Variety) %>%
  summarise(
    N = n(),
    Mean = mean(Yield),
    SD = sd(Yield),
    .groups = "drop"
  )
```

## Model Fitting

We fit a one-way ANOVA using `aov()`:

```{r model}
mod <- aov(Yield ~ Variety, data = crd_data)
summary(mod)
```

The F-test assesses whether at least one variety mean differs from the others.

## Assumption Checking

ANOVA assumes that residuals are normally distributed and that variances
are homogeneous across groups.

```{r assumptions, fig.width = 10, fig.height = 4}
check_assumptions(mod, data = crd_data, group = "Variety")
```

We can also check normality per group:

```{r shapiro-group}
crd_data %>%
  group_by(Variety) %>%
  summarise(
    Shapiro_p = shapiro.test(Yield)$p.value,
    .groups = "drop"
  )
```

## Post-hoc Comparisons

If the F-test is significant, we compare varieties pairwise using Tukey's HSD:

```{r tukey}
TukeyHSD(mod)
```

```{r tukey-plot}
plot(TukeyHSD(mod), las = 1, col = "brown")
```

### Estimated Marginal Means

The `emmeans` package provides estimated marginal means and pairwise contrasts:

```{r emmeans}
library(emmeans)

emm <- emmeans(mod, specs = "Variety")
emm
```

```{r emmeans-pairs}
pairs(emm)
```

```{r emmeans-plot, fig.width = 7, fig.height = 4}
plot(emm, comparisons = TRUE) +
  theme_bw() +
  labs(y = "", x = "Estimated marginal mean")
```

## Conclusion

The CRD analysis provides a straightforward way to compare treatment means
when experimental units are homogeneous. The key outputs are:

1. **F-test** from the ANOVA table to detect overall differences.
2. **Diagnostic plots** to verify model assumptions.
3. **Tukey HSD** or **emmeans** contrasts for pairwise comparisons.

When experimental units are not homogeneous, consider using a
[Randomized Complete Block Design](02-rcbd.html) instead.
