---
title: "Randomized Complete Block Design (RCBD)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Randomized Complete Block Design (RCBD)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## When to Use

The **Randomized Complete Block Design (RCBD)** is used when:

- There is **one treatment factor** of interest.
- There is a **known source of variability** (e.g., soil gradient, time, lab bench)
  that can be controlled by grouping units into **blocks**.
- Each block contains **all treatments** exactly once.

RCBD reduces experimental error by accounting for block-to-block variability,
making it more efficient than a CRD when conditions are not uniform.

## The Design

The statistical model is:

$$Y_{ij} = \mu + \beta_j + \tau_i + \varepsilon_{ij}$$

where $\beta_j$ is the effect of block $j$ and $\tau_i$ is the treatment effect.
Block effects are included to absorb known variability, but we are primarily
interested in testing treatment differences.

## Data

We use a dataset with six fertilizer levels (X25 through X150) applied across
four blocks.

```{r data}
library(agrideshr)
data(rcbd_data)
str(rcbd_data)
head(rcbd_data, 12)
```

## Exploratory Visualization

```{r boxplot}
library(ggplot2)

ggplot(rcbd_data, aes(x = Treatment, y = Yield, fill = Treatment)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, size = 2) +
  theme_bw() +
  labs(title = "Yield by Fertilizer Level", y = "Yield") +
  theme(legend.position = "none")
```

Treatment response by block:

```{r interaction}
ggplot(rcbd_data, aes(x = Treatment, y = Yield, colour = Block, group = Block)) +
  geom_point(size = 2) +
  stat_summary(fun = mean, geom = "line") +
  theme_bw() +
  labs(title = "Treatment Response by Block")
```

## Model Fitting

In RCBD, the block is included as a factor in the model but is not the focus
of inference:

```{r model}
mod <- aov(Yield ~ Block + Treatment, data = rcbd_data)
summary(mod)
```

The F-test for `Treatment` tells us whether fertilizer levels differ after
accounting for block effects.

## Assumption Checking

```{r assumptions, fig.width = 10, fig.height = 4}
check_assumptions(mod, data = rcbd_data, group = "Treatment")
```

## Post-hoc Comparisons

We compare treatments (not blocks) using Tukey's HSD:

```{r tukey}
TukeyHSD(mod, which = "Treatment")
```

```{r tukey-plot}
plot(TukeyHSD(mod, which = "Treatment"), las = 1, col = "brown")
```

### Estimated Marginal Means

```{r emmeans}
library(emmeans)

emm <- emmeans(mod, specs = "Treatment")
emm
```

```{r emmeans-pairs}
pairs(emm)
```

```{r emmeans-plot, fig.width = 7, fig.height = 4}
plot(emm, comparisons = TRUE) +
  theme_bw() +
  labs(y = "", x = "Estimated marginal mean")
```

### Block-by-Treatment Interaction Plot

This plot helps visualise whether the treatment effect is consistent across blocks:

```{r emmip}
emmip(mod, Block ~ Treatment) +
  theme_bw() +
  labs(title = "Block x Treatment Interaction Plot")
```

## Conclusion

The RCBD is the workhorse of agricultural experimentation. By blocking on a
known source of variability, it provides a more precise estimate of treatment
effects than a CRD. Key outputs:

1. **ANOVA table** with block and treatment effects separated.
2. **Tukey HSD** for pairwise treatment comparisons.
3. **EMM interaction plots** to check consistency across blocks.

When there are **two sources of heterogeneity** (e.g., rows and columns),
consider a [Latin Square Design](03-latin-square.html).
