---
title: "Latin Square Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Latin Square Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## When to Use

The **Latin Square Design (LSD)** extends the RCBD by controlling for **two
sources of variability** simultaneously. Use it when:

- You have **one treatment factor** of interest.
- There are **two blocking factors** (e.g., rows and columns in a field,
  days and operators in a lab).
- The number of treatments equals the number of levels in each blocking factor
  (producing a square layout).

## The Design

Each treatment appears exactly once in every row and every column. The model is:

$$Y_{ijk} = \mu + \rho_i + \gamma_j + \tau_k + \varepsilon_{ijk}$$

where $\rho_i$ is the row effect, $\gamma_j$ is the column effect, and
$\tau_k$ is the treatment effect.

## Data

We use a 4x4 Latin square dataset with four treatments, four rows, and four
columns.

```{r data}
library(agrideshr)
data(lsd_data)
str(lsd_data)
lsd_data
```

## Exploratory Visualization

```{r scatter}
library(ggplot2)

ggplot(lsd_data, aes(x = Treatment, y = Yield, colour = Row, shape = Column)) +
  geom_point(size = 3) +
  ylim(0, NA) +
  theme_bw() +
  labs(title = "Yield by Treatment (Row and Column indicated)")
```

## Model Fitting

Both row and column effects are included in the model:

```{r model}
mod <- aov(Yield ~ Row + Column + Treatment, data = lsd_data)
summary(mod)
```

The F-test for `Treatment` assesses treatment differences after removing row
and column variability.

## Assumption Checking

```{r assumptions, fig.width = 10, fig.height = 4}
check_assumptions(mod, data = lsd_data, group = "Treatment")
```

## Post-hoc Comparisons

We compare treatments only (rows and columns are nuisance factors):

```{r tukey}
TukeyHSD(mod, which = "Treatment")
```

```{r tukey-plot}
plot(TukeyHSD(mod, which = "Treatment"), las = 1, col = "brown")
```

### Estimated Marginal Means

```{r emmeans}
library(emmeans)

emm <- emmeans(mod, specs = "Treatment")
emm
```

```{r emmeans-pairs}
pairs(emm)
```

```{r emmeans-plot, fig.width = 7, fig.height = 4}
plot(emm, comparisons = TRUE) +
  theme_bw() +
  labs(y = "", x = "Estimated marginal mean")
```

### Interaction Plots

Visualise treatment effects across rows and columns:

```{r emmip-row}
emmip(mod, Column ~ Treatment | Row) +
  theme_bw() +
  labs(title = "Column effect across Treatments by Row")
```

```{r emmip-col}
emmip(mod, Row ~ Treatment | Column) +
  theme_bw() +
  labs(title = "Row effect across Treatments by Column")
```

## Conclusion

The Latin Square Design controls for two sources of heterogeneity simultaneously,
making it highly efficient when the experimental layout naturally forms a grid.
Limitations include:

- The number of treatments must equal the number of row and column levels.
- No interaction terms can be estimated (df are used for blocking).

When you need to study **two treatment factors** and their interaction,
consider a [Factorial Design](04-factorial.html).
