---
title: "Factorial Design (Two-Way ANOVA)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factorial Design (Two-Way ANOVA)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## When to Use

A **Factorial Design** is used when:

- You want to study **two or more treatment factors** simultaneously.
- You are interested in both **main effects** and **interactions** between factors.
- All combinations of factor levels are tested.

Factorial designs are more efficient than running separate one-factor experiments
because they provide information about interactions at no extra cost.

## The Design

For a two-factor factorial, the model is:

$$Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \varepsilon_{ijk}$$

where $\alpha_i$ is the effect of factor A (e.g., Temperature), $\beta_j$ is the
effect of factor B (e.g., Fertilizer), and $(\alpha\beta)_{ij}$ is their interaction.

## Data

We use a dataset crossing three temperature levels with three fertilizer types,
with four replicates per combination.

```{r data}
library(agrideshr)
data(factorial_data)
str(factorial_data)
head(factorial_data, 12)
```

## Exploratory Visualization

```{r interaction-plot}
library(ggplot2)

ggplot(factorial_data, aes(x = Temperature, y = Yield, colour = Fertilizer,
                            group = Fertilizer)) +
  geom_point(size = 2, position = position_dodge(0.2)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  theme_bw() +
  labs(title = "Interaction Plot: Temperature x Fertilizer")
```

```{r boxplot}
ggplot(factorial_data, aes(x = Temperature, y = Yield, fill = Fertilizer)) +
  geom_boxplot(alpha = 0.7) +
  theme_bw() +
  labs(title = "Yield by Temperature and Fertilizer")
```

## Model Fitting

### Additive Model (No Interaction)

```{r additive}
mod_add <- aov(Yield ~ Temperature + Fertilizer, data = factorial_data)
summary(mod_add)
```

### Full Model (With Interaction)

```{r full}
mod_full <- aov(Yield ~ Temperature * Fertilizer, data = factorial_data)
summary(mod_full)
```

If the interaction term is not significant, the additive model is preferred for
its simplicity. If significant, the interaction must be interpreted.

## Assumption Checking

```{r assumptions, fig.width = 10, fig.height = 4}
check_assumptions(mod_add, data = factorial_data, group = "Temperature")
```

## Post-hoc Comparisons

### By Fertilizer

```{r tukey-fert}
TukeyHSD(mod_add, which = "Fertilizer")
```

```{r tukey-fert-plot}
plot(TukeyHSD(mod_add, which = "Fertilizer"), las = 1, col = "brown")
```

### By Temperature

```{r tukey-temp}
TukeyHSD(mod_add, which = "Temperature")
```

```{r tukey-temp-plot}
plot(TukeyHSD(mod_add, which = "Temperature"), las = 1, col = "brown")
```

### Estimated Marginal Means

```{r emmeans-fert}
library(emmeans)

emm_fert <- emmeans(mod_add, specs = "Fertilizer")
emm_fert
pairs(emm_fert)
```

```{r emmeans-temp}
emm_temp <- emmeans(mod_add, specs = "Temperature")
emm_temp
pairs(emm_temp)
```

```{r emmeans-plots, fig.width = 10, fig.height = 4}
p1 <- plot(emm_fert, comparisons = TRUE) +
  theme_bw() + labs(y = "", x = "EMM: Fertilizer")
p2 <- plot(emm_temp, comparisons = TRUE) +
  theme_bw() + labs(y = "", x = "EMM: Temperature")
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

### Conditional Contrasts (If Interaction Is Present)

When the interaction is significant, compare one factor at each level of the other:

```{r conditional, eval = FALSE}
# Temperature comparisons within each Fertilizer level
emmeans(mod_full, pairwise ~ Temperature | Fertilizer)

# Fertilizer comparisons within each Temperature level
emmeans(mod_full, pairwise ~ Fertilizer | Temperature)
```

### Interaction Plot (EMM)

```{r emmip}
emmip(mod_add, Fertilizer ~ Temperature) +
  theme_bw() +
  labs(title = "Estimated Marginal Means: Fertilizer x Temperature")
```

## Conclusion

Factorial designs allow efficient study of multiple factors and their interactions.
The analysis workflow:

1. Fit both additive and interaction models.
2. Test whether the interaction is significant.
3. If **no interaction**: interpret main effects separately.
4. If **interaction present**: use conditional contrasts to compare one factor
   within levels of the other.

When factors have a **nested structure** (e.g., sub-treatments applied within
main treatments), consider a [Split-Plot Design](05-split-plot.html).
