---
title: "Split-Plot Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Split-Plot Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## When to Use

The **Split-Plot Design** is used when:

- You have **two treatment factors**, but one is harder (or more expensive)
  to change than the other.
- The **whole-plot factor** is applied to larger units, and the **sub-plot factor**
  is applied within those units.
- Randomisation is **restricted**: the whole-plot factor is randomised first,
  then the sub-plot factor is randomised within each whole plot.

Common examples: irrigation method (whole-plot) x fertilizer type (sub-plot),
or tillage system (whole-plot) x crop variety (sub-plot).

## The Design

The split-plot design has **two error terms**:

$$Y_{ijk} = \mu + R_k + A_i + \delta_{ik} + B_j + (AB)_{ij} + \varepsilon_{ijk}$$

where:

- $R_k$ = replicate (block) effect
- $A_i$ = whole-plot treatment effect
- $\delta_{ik} \sim N(0, \sigma^2_\delta)$ = whole-plot error
- $B_j$ = sub-plot treatment effect
- $(AB)_{ij}$ = interaction
- $\varepsilon_{ijk} \sim N(0, \sigma^2)$ = sub-plot error

This requires a **mixed model** because the whole-plot error is a random effect.

## Data

We use a dataset with 6 whole-plot treatments, 3 sub-plot treatments, and
6 replicates.

```{r data}
library(agrideshr)
data(split_plot_data)
str(split_plot_data)
head(split_plot_data, 12)
```

## Exploratory Visualization

```{r facet-plot}
library(ggplot2)

ggplot(split_plot_data, aes(x = WholePlot, y = Yield, colour = SubPlot,
                             group = SubPlot)) +
  geom_line() +
  facet_wrap(~ Replicate) +
  theme_bw() +
  labs(title = "Yield by Whole-Plot and Sub-Plot Treatment (by Replicate)")
```

```{r boxplot}
ggplot(split_plot_data, aes(x = WholePlot, y = Yield, fill = SubPlot)) +
  geom_boxplot(alpha = 0.7) +
  theme_bw() +
  labs(title = "Yield by Whole-Plot and Sub-Plot Treatment")
```

## Model Fitting

Unlike the previous designs, split-plot requires a **linear mixed model**
using `lmer()` from the `lmerTest` package. The term `(1 | Replicate:WholePlot)`
captures the whole-plot error:

```{r model}
library(lmerTest)

fit <- lmer(Yield ~ Replicate + WholePlot * SubPlot + (1 | Replicate:WholePlot),
            data = split_plot_data)
```

### ANOVA Table

```{r anova}
anova(fit)
```

The F-tests use appropriate denominator degrees of freedom:

- **WholePlot** is tested against the whole-plot error.
- **SubPlot** and **WholePlot:SubPlot** are tested against the sub-plot (residual) error.

### Variance Components

```{r varcomp}
VarCorr(fit)
```

## Assumption Checking

For mixed models, we check residuals at the sub-plot level:

```{r assumptions, fig.width = 8, fig.height = 4}
resid_vals <- residuals(fit)
fitted_vals <- fitted(fit)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(fitted_vals, resid_vals,
     xlab = "Fitted", ylab = "Residuals",
     main = "Residuals vs Fitted", pch = 16, col = "steelblue")
abline(h = 0, lty = 2, col = "red")

qqnorm(resid_vals, main = "Normal Q-Q Plot", pch = 16, col = "steelblue")
qqline(resid_vals, col = "red", lwd = 2)
```

```{r shapiro}
shapiro.test(resid_vals)
```

## Post-hoc Comparisons

### Estimated Marginal Means

#### Whole-plot treatment within each sub-plot level

```{r emm-a-by-b}
library(emmeans)

emm_a <- emmeans(fit, ~ WholePlot | SubPlot)
pairs(emm_a)
```

```{r emm-a-plot, fig.width = 8, fig.height = 5}
plot(emm_a, comparisons = TRUE) +
  theme_bw() +
  labs(y = "", x = "Estimated marginal mean")
```

#### Sub-plot treatment within each whole-plot level

```{r emm-b-by-a}
emm_b <- emmeans(fit, ~ SubPlot | WholePlot)
pairs(emm_b)
```

```{r emm-b-plot, fig.width = 8, fig.height = 5}
plot(emm_b, comparisons = TRUE) +
  theme_bw() +
  labs(y = "", x = "Estimated marginal mean")
```

### Interaction Plots

```{r emmip-1}
emmip(fit, WholePlot ~ SubPlot) +
  theme_bw() +
  labs(title = "Whole-Plot x Sub-Plot Interaction")
```

```{r emmip-2}
emmip(fit, SubPlot ~ WholePlot) +
  theme_bw() +
  labs(title = "Sub-Plot x Whole-Plot Interaction")
```

## Conclusion

The split-plot design is essential when practical constraints prevent full
randomisation of all treatment combinations. Key points:

1. **Mixed model** is required (`lmer`, not `aov`) to correctly partition
   the two error terms.
2. The whole-plot factor is tested with **less precision** (fewer effective
   replicates) than the sub-plot factor.
3. **emmeans** with conditional contrasts (`~ A | B`) is the appropriate
   way to make pairwise comparisons.

## Design Summary

| Design | Vignette | Factors | Blocking | Model |
|--------|----------|---------|----------|-------|
| CRD | [01](01-crd.html) | 1 | None | `aov(Y ~ A)` |
| RCBD | [02](02-rcbd.html) | 1 + Block | 1 blocking factor | `aov(Y ~ Block + A)` |
| Latin Square | [03](03-latin-square.html) | 1 + Row + Col | 2 blocking factors | `aov(Y ~ Row + Col + A)` |
| Factorial | [04](04-factorial.html) | 2+ crossed | Optional | `aov(Y ~ A * B)` |
| Split-Plot | [05](05-split-plot.html) | 2 nested | Replicates | `lmer(Y ~ A*B + (1|Rep:A))` |
